Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-08-10 Manuscript/MPESubmission/Revision/response.tex

See also appendix B section "Differences between the ``true'' and the ``best'' trees."
% NC: Needs a little more explanation, and a line about what analysis you did in appendix b. So something along the lines of "We compare best trees to missing data trees, but could also compare true trees to the missing data trees. We do not do this because... However, results of analyses comparing the two can be found in appendix B. This makes little difference to our overall results." That might need to be better integrated with the paragraph(s) before.

% -----------------
% Comment 7
% -----------------
\item{\textcolor{blue}{\textbf{Methodology - Starting tree:} The authors should at least include some analysis in which no starting tree is given, the max. likelihood 'best' tree is given, and / or the 'best' tree is given but with some perturbation.
I fear that one of the main results may be biased due to this one methodological detail - a strong prior on tree topology.
The justification given was 'to speed up the Bayesian analysis' which does not seem adequate, given these datasets are small.}}

%%%%
We take on board the coment and agree that this would be a bad bias if it was the case. However we ran extra analysis and there is no bias.
%%%%

We performed an extra analysis on a subset of five trees with the three missing data parameters at 0\%, 10\%, 25\%, 50\% and 75\% for 20 chains. 
We compared the topology of these trees in Bayesian inference when using the true tree as a starting tree or not using a starting tree (i.e. using the default in MrBayes which is using a random tree).

% NC: Yes but the reviewer is asking for something different - they ask you to have no starting tree. If this is impossible in MrBayes then say so, and say you used a random tree instead. TG: Maybe there's a phylogeny confusion here. In both ML or Bayesian, you need to start from a tree, then calculate the likelihood of all the parameters, then change the tree, calculate again, etc...
We found no difference in topologies between both the Bayesian inferences starting with the true tree or a random tree.
Therefore, we do not believe our results are merely the result of using a specific starting tree topology. %strong priors on the tree topology. TG: That's the big difference that the PLoS ONE reviewer also didn't got. A starting tree is not a Bayesian prior (in P(A|B) = P(B|A)*P(A)/P(B; P(A) is the prior). It's just a user fixed starting point for the MCMC walk rather than a random one.
% NC: The reviewer also says that saving time is not a good justification. It'd be worth explaining here how long it actually takes even though these datasets are small.
%
% %%%%%%%%%%%%%%%%%%%
% TO DISCUSS @@@
% %%%%%%%%%%%%%%%%%%%
%
% TG: about that: it using a random tree or the true tree doesn't change the timing significantly. The whole idea was suggested by Rob or Gavin (don't remember exactly) at lunch time when I was visiting Gavin in Sheffield. To speed up the search in the tree space they suggested I already start from a sensible tree in the tree space, that was meant to speed up the analysis.
%
These results are available in the supplementary materials Appendix A (figure A2 and tables A1 and A2) and we also added a note to the manuscript:

@@

We changed the axes of the graphs showing Normalised Triplets metric to show values below 0 throughout the manuscript and the appendices.

% -----------------
% Comment 14
% -----------------
\item{\textcolor{blue}{\textbf{Table caption:} Table caption should include a line about 'pooling across missing data schemes' to clarify what is presented.}}

%TG: WUT? Didn't understood that one at all.
% NC: The reviewer means that you're presenting results from across all your different parameters, and all your different amounts of missing data in the table. You need to clarify this, especially as in figures you separate these out. Pooling just means combining. I noticed this on my re-read just now too.

% TG: Add that to the caption
% Pooled the data across the different missing data parameters

% %%%%%%%%%%%%%%%%%%%
% TO DISCUSS @@@
% %%%%%%%%%%%%%%%%%%%

@@

(1) - We thank the reviewer for this comment and now emphasize both the novel aspects of our study and the questions we are actually testing (see reviewer 1 comment 4 above).

(2) - the second point underlined by the reviewer in this specific comment strikes use as really interesting. However, we considered this study to already be long and complex. We therefore think that adding an extra question to this study (i.e. what is the effect of missing data on the morphological substitution rate parameter estimates?) might lose the reader and make an already long paper even longer.
Nonetheless, we added the result of these analyses to \textbf{Appendix-A Morphological rates estimations results}.
We link to this analysis in the methods as follows:

"The detailed MrBayes parameters are available in Appendix A along with details on the $\alpha$ parameter estimation." lines 336-338 @@@
See also Appendix A "Morphological rates estimations results".

We did it
but we put in appendix
because it's too long.

% TG: what do you think? I don't want to attract the reviewer's attention too much (i.e. that's another shit load of analysis to do to do it properly) and certainly not the reader's attention either (it's not relevant to the whole paper) but sticking that to the supplementary kind of does what the reviewers wants to know.

% NC: I'm a bit confused - did you do what they wanted or not? If not then the whole "this will make the paper longer" argument holds. If you did do what they wanted then it might need a little more explanation somewhere. You can't just throw random stuff into the appendices for no reason.
%
% %%%%%%%%%%%%%%%%%%%
% TO DISCUSS @@@
% %%%%%%%%%%%%%%%%%%%
%
% TG: Ok, maybe we should remove it. I've done the analysis because it's interesting and easy to do but maybe we can just add the results to this response, not to the revised manuscript?

@@

% -----------------
% Specific Comment 17
% -----------------
\item{\textcolor{blue}{\textbf{line 173:} streamline to "The substitution rates were selected from a gamma distribution with…" or something similar. }}

We modified the sentence as suggested.

% -----------------
% Specific Comment 18
% -----------------
\item{\textcolor{blue}{\textbf{line 173 to 176:} need some justification for why the substitution rate with alpha shape of 0.5  will avoid homoplasy. Yang 1996 is cited but clarify the justification of what is 'too much' homoplasy. }}

Homoplasy is a difficult concept, not really a value so we went for...
We changed the sentence to:
"In practice, a value of $\alpha$ $<$ 1 decreases the number of sites with high substitution rates, thus reducing homoplasic sites and increasing the phylogenetic signal (Hassanin et al., 1998; Estoup et al., 2002).%\citep{Hassanin1998611,EstoupHomoplasy}.
Also, we chose this low value to be consistent with our protocol for simulating morphological characters (see below)." (line 189-192@@@).

% NC: I think what is needed here is a definition of what is "too much". Is it 50%? 20%? etc. Too much is arbitrary.
% %%%%%%%%%%%%%%%%%%%
% TO DISCUSS @@@
% %%%%%%%%%%%%%%%%%%%
% TG: That's why I removed the "too much". I'm actually not sure how Homoplasy is measured (as in units of homoplasy). Wagner did a lot of work on it but he usually refers to homoplasy through proxies like characters retention, etc...

@@

We modified the results section as follows:
"Using both Normalised Robinson-Foulds and Normalised Triplets metrics from the Bayesian consensus trees, the parameter combination with no missing data (i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $N_{C}$ = 100) is always the most dissimilar to all the other parameter combinations (thin deep blue line at the base of Figure 6).
The Normalised Robinson-Foulds metric (median Bhattacharrya coefficient = 0.79; blue regions in Figure 6A), however, displays more dissimilarities than the Normalised Triplets metric (median Bhattacharrya coefficient = 0.81; blue regions in Figure 6B).
The orange upper triangle in Figure 6A shows a high probability of overlap of the Normalised Robinson-Foulds metric for the trees with the $M_{L}$ parameter $\geq$ 50\% (Figure 6A).
% There is no additional effect of $M_{F}$ and $N_{C}$, regardless of the amount of missing data in these parameters (Figure 6A).
% Additionally, when using the Normalised Robinson-Foulds metric, once $M_{L}$ $\geq$ 50\%, there is no additional effect of $M_{F}$ and $N_{C}$, regardless of the amount of missing data in these parameters (Figure 6A).
% NC: This sentence and the one before it seem to be the same thing? TG: Yep, I modified to:
Once $M_{L}$ $\geq$ 50\%, there is no additional effect of $M_{F}$ and $N_{C}$, regardless of the amount of missing data in these parameters (Figure 6A).
Likewise, once $N_{C}$ $<$ 50, there is no additional effect of $M_{L}$ and $M_{F}$ as denoted by the high probability of Normalised Robinson-Foulds metric overlap (horizontal orange stripes between the blue regions Figure 6A).
In figure 5 for the Normalised Robinson-Foulds metric, this can be interpreted as the overlap between the distributions once $M_L$=50\%." lines 480-497@@@

% -----------------
% Specific Comment 58
% -----------------
\item{\textcolor{blue}{\textbf{Table 1:} The wording of the caption is off - the second paragraph is awkward. 
Are these values the probabilities of overlap between the distributions of the 'best' tree versus trees from each inference method pooled across all combinations of missing data? That is what is described in methods line 397 but it doesn't say so in the caption.
Two things:\\
1)  Comparing to the tables in Appendix C, the distribution of probabilities of overlap between max. likelihood and Bayesian trees are higher in most missing data schemes in tables in the Appendices than the overall values given in Table 1. How could the max. likelihood and Bayesian trees be so different?\\
% TG: we really apologize but we think there's an error. We don't know where their coming from but it's probably our fault because of mislabelling. Hopefully this is fixed.

% Check again table. Difference between tables and figures.

2)  Pooling across all missing data schemes seems to wash out the effect of tree-inference method. I am not sure what a better way to present these would be, but perhaps some kind of matrix where the cells are the joint probabilities of overlap of each method with each missing data category, and then the marginal probabilities would be the probabilities of overlap for each method, averaged over missing data categories, as well as the marginal probabilities of overlap for each missing data category, averaged across methods.  e.g, RF distance:\\
Missing data:   0   25  50  75  Marginal \\
BI  1   0.9 0.8 0.75    0.863 \\
ML  1   0.9 0.75    0.25    0.725 \\
Marginal    1   0.9 0.78    0.5  \\
This may not be feasible in the text due to space limitations, maybe tables like these could be added to an appendix.}}

@@

% %%%%%%%%%%%%%%%%%%%
% TO DISCUSS @@@
% %%%%%%%%%%%%%%%%%%%

2) %TG: not entirely sure what the reviewer means by marginal probabilities. I made to extra summarizing tables in the supplementaries that may be actually easier to read than most of the others so we might want to add them to the main text.
We added two extra tables in the Appendix C (Table C8 and C9) that summarized the median Bhattacharyya Coefficients within each methods and each parameters individually and combined (Table C8; as illustrated in Figure 2) as well as the median Bhattacharyya Coefficients within each methods and each parameters individually and combined (Table C9; as illustrated in Figure 3).
% TG: If that's actually what the reviewer 1 wanted, I think we can propose to add it to the main text rather than the appendices. They can always decide to move it back to the appendices if it doesn't bring much.

% NC: https://en.wikipedia.org/wiki/Marginal_distribution
% NC: Marginal probabilities are basically the probability of one thing given you're controlling for another thing. So they are looking to see the effect of tree inference method independent of parameter differences. Otherwise they suggest you are hiding important differences via this summarizing over all parameter combinations. 
% NC: I think you can argue that yes, table 1 hides some important effects, but that it was difficult to summarize these results and more in depth tables can be seen in the appendix.

% And fix this one with adding the tables.
% Marginal probabilities checking. Do it?

@@

% -----------------
% Specific Comment 59
% -----------------
\item{\textcolor{blue}{\textbf{Appendix C:} Tables 1-3 repeats 'of the'. Should label the first two rows as 'Best maximum likelihood tree' and 'Best Bayesian inference tree'}}

We removed the repetition in the caption.
% If we understand the second point correctly, we have not added "best" to these as the Maximum Likelihood tree is by definition the tree with the highest likelihood, and the Bayesian consensus is the consensus tree of the Bayesian inference trees. So adding the word best here is not entirely accurate.

Change caption in the "best" ML tree and "best" Bay con tree to blablabla.

### 2015-08-10 Manuscript/MPESubmission/Revision/TEM_revised_manuscript.tex
Comments on revisions

\noindent
\textbf{Table 1:} \hl{Bhattacharyya Coefficients of the pairwise method comparisons.}
\hlEach line summarizes the probabilities of overlap between the distributions of the ``best'' tree versus trees from each inference method (Maximum Likelihood; Bayesian consensus; Maximum Likelihood Bootstraps and Bayesian posterior trees) pooled across all combinations of missing data parameter values, using the Normalised Robins-Foulds (RF) and Triplets (Tr) metrics. 
Values highlighted in bold are the extreme values of high or low probability of overlap between two methods. If two methods have a high probability of overlap, they have a similar ability to recover the ``correct'' tree topology.}\\

>>>>

\noindent
\textbf{Table 1:} \hl{Bhattacharyya Coefficients of the pairwise method comparisons.}
\hl{Each line summarizes the probabilities of overlap between the distributions of the ``best'' tree versus trees from each inference method (Maximum Likelihood; Bayesian consensus; Maximum Likelihood Bootstraps and Bayesian posterior trees) pooled across all combinations of missing data parameter values, using the Normalised Robins-Foulds (RF) and Triplets (Tr) metrics. 
Values highlighted in bold are the extreme values of high or low probability of overlap between two methods. If two methods have a high probability of overlap, they have a similar ability to recover the ``correct'' tree topology.}\\
